Unshuffling Permutations

### 2015-09-18 article.tex
Improvements

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$
are specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the
$i$-th letter of any permutation $\pi$.
For any nonnegative integer $n$, we write $\nearrow_{n}$
(resp. $\searrow_{n}$)
for the permutation $1 2 \ldots n$ (resp. $n (n-1) \ldots 1$).
If $\pi$ is a permutation of $S_n$,
we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is the
permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

If $u$ is a word of integers without multiple
occurrences of a same letter, $\STD(u)$ is the {\em standarized} of $u$,
that is the unique permutation of the same size as $u$ such that for all
$i, j \in [|u|]$, $u_i < u_j$ if and only if $\STD(u)_i < \STD(u)_j$. In
particular, the image of the map $\STD$ is the set $S$ of all
permutations. Two words $u$ and $v$ having the same standarized are
{\em order-isomorphic}. If $\sigma$ is a permutation, there is an
{\em occurrence} of (the {\em pattern}) $\sigma$ in $\pi$ if there is a
set $P$ of indexes of letters of $\pi$ such that $\sigma$ and $\pi_{|P}$
are order-isomorphic. When $\pi$ does not admit any occurrence of
$\sigma$, $\pi$ {\em avoids} $\sigma$.

For a graph $G$, we denote $\VS{G}$ as the set of vertices and
$\ES{G}$ as the set of edges.
Let $u = u_1 \, u_2 \,\ldots\, u_n$ be a word.
The \emph{clique graph associated to $u$}, denoted $K_u$,
is the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$.
The structure of this underlying graph is linear,
\emph{i.e.}, the set of vertices is equipped with a natural total order
$<$ defined by $u_i < u_j$ if and only if $i < j$.
For a linear graph, we write $(u_i, u_j)$ instead of $\{u_i, u_j\}$
for an edge if it is clear from the context that $i < j$.
Recall that two edges of a graph are \emph{independent} if they do not share
a common vertex, and that a \emph{matching} $\mathcal{M}$ in $G$ is a set of
pairwise independent edges.
A matching is \emph{perfect} if it covers all the vertices of the graph.
In case the set of vertices is equipped with a total order,
a matching $\mathcal{M}$ is said to be \emph{containment-free}
if there do not exist (independent) edges $(u_i, u_j)$ and
$(u_k, u_\ell)$ in $\mathcal{M}$ such that
$u_i < u_k < u_\ell < u_j$ or  $u_k < u_i < u_j < u_\ell$.

@@

        Finally, observe that any permutation $\pi$ avoiding $213$ and
        $231$ necessarily starts by the smallest possible letter or the
        greatest possible letter. This property is then true for the
        suffix of $\pi$ obtained by deleting its first letter,
        and so on for all of its suffixes. Thus, by
        replacing each letter $a$ of $\pi$ by $0$ (resp. $1$) if $a$ has the
        role of a smallest (resp. greatest) letter, one obtains a binary
        word $u$ such that $\BINTOPERM(u) = \pi$. Hence, all permutations
        avoiding $213$ and $231$ are in the image of $\BINTOPERM$.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
        Since $u$ is a square binary word, there is a binary word $v$
        such that $u \in v \shuffle v$. Then, there are two disjoint
        sets $P$ and $Q$ of positions of letters of $u$ such that
        $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
        words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
        same standarized $\rho$. Hence, and by definition of the shuffle
        product of permutations, $\BINTOPERM(u)$ appears in
        $\rho \SHUFFLE \rho$, showing that $\BINTOPERM(u)$ is a square
        permutation.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_permutation_to_square_binary}]
        Let $\pi$ be a square permutation avoiding $213$ and $231$. By
        Claim~\ref{claim:binary_to_permutation_avoiding}, $\pi$ is in
        the image of $\BINTOPERM$ and hence, $u = \PERMTOBIN(\pi)$ is a
        well-defined binary word. Since $\pi$ is a square permutation,
        there are two disjoint sets $P_1$ and $P_2$ of indexes of letters
        of $\pi$ such that $\pi_{|P_1}$ and $\pi_{|P_2}$ are
        order-isomorphic. This implies, by the definitions of $\BINTOPERM$
        and $\PERMTOBIN$, that $u_{|P_1} = u_{|P_2}$, showing that $u$
        is a square binary word.
        \qed
    \end{proof}
\end{proof}
\medskip

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\rho$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\rho}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \rho$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\rho^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Lemma~\ref{lem:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\rho$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\rho \SHUFFLE \rho$. Now, by Lemma~\ref{lem:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\rho \SHUFFLE \rho) =
        \phi_j(\rho) \SHUFFLE \phi_j(\rho).
    \end{equation}
    Then, since $\pi$ appears in $\rho \SHUFFLE \rho$, $\phi_j(\pi)$
    appears in $\phi_j(\rho \SHUFFLE \rho)$ and appears also in
    $\phi_j(\rho) \SHUFFLE \phi_j(\rho)$. This shows that
    $\phi_j(\rho)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}
\medskip

@@

This section is devoted to proving hardness of recognizing square permutations.
In the same way as happens with words, we shall use a linear graph framework
where deciding whether a permutation is a square reduces to computing some
specific matching in the associated linear graph \cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}.
It reads as follows.

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item
    The permutation $\pi$ is a square.
    \item
    There exists a perfect containment-free mathing
    $\mathcal{M}$ of $G_\pi$ such that
    for any two disctinct edges
    $(a, b)$ and $(a', b')$ in $\mathcal{M}$
    we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

  (\emph{1} $\Rightarrow$ \emph{2}).
  Since $\pi$ is a quare, it admits a square root $\sigma$. Then, by
  definition, there
  are two sets $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$ and
  $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
  letters of $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
  order-isomorphic to $\sigma$.
  %Let $\sigma$ be a permutation such that $\pi$ is the union of two
  %patterns that are order-isomorphic to $\sigma$.
  Let $2n = |\pi|$ (and hence $|\sigma| = n$).
  %Fix the occurrences in $\pi$ of the two disjoint patterns that
  %are order-isomorphic to $\sigma$ and
  %write
  %$I^1 = \{i^1_1, i^1_2, \ldots, i^1_n\}$,
  %$i^1_1 < i^1_2 < \ldots < i^1_n$, for the positions in $\pi$
  %of the first pattern order-ismorphic to $\sigma$ and
  %$I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$,
  %$i^2_1 < i^2_2 < \ldots < i^2_n$, for the positions in $\pi$
  %of the second pattern order-ismorphic to $\sigma$.
  Let $G_\pi$ be the linear clique associated to $\pi$ with
  vertices $\VS{G_\pi} =
  \{i^1_j : 1 \leq j \leq n\} \cup \{i^2_j : 1 \leq j \leq n\}$.
  It is easily seen that
  $\mathcal{M} = \{\{i^1_j, i^2_j\} : 1 \leq j \leq n\}$
  is a subset of $\ES{G_\pi}$.
  Furthermore, $\mathcal{M}$ is a perfect matching since
  $I^1 \cap I^2 = \emptyset$,
  and $I^1 \cup I^2 = \{1, 2,  \ldots 2n\}$.
  It is also containment-free.
  Indeed, if it is not the case then there would exist two edges
  $e = \{i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$, $j < k$,
  in $\mathcal{M}$ such that
  $i^1_j < i^1_k < i^2_k < i^2_j$ or
  $i^1_j < i^2_k < i^1_k < i^2_j$.
  This is a contradiction since
  $i^2_k > i^2_j$ if $k > j$.
  Finally, for any two distinct edges
  $e = \{\i^1_j, i^2_j\}$ and
  $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
  we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
  $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing
  in both cases two elements (at position $j$ and $k$)
  in two patterns that are order-isomorphic to $\sigma$.

  (\emph{2} $\Rightarrow$ \emph{1}).
  Let $I^1 = \{i^1_j : \exists \{i_j, i_k\} \in \mathcal{M}
  \;\text{with}\; j < k\}$, and
  $I^2 = \{i^2_k : \exists \{i_j, i_k\} \in \mathcal{M}
  \;\text{with}\; j < k\}$.
  Informally, $I^1$ (resp. $I^2$) is the set of the left (resp. right) endpoints
  of the edges in $\mathcal{M}$ assuming the positions $1, 2, \ldots, 2$
  are drawn along an horizontal line.
  Without loss of generality,
  assume $i^1_1 < i^1_2 < \ldots < i^1_n$.
  % and
  % $I^2 = \{i^2_1, i^2_2, \ldots, i^2_n\}$.
  Let
  $p^1 = \pi(i^1_1) \pi(i^1_2) \ldots \pi(i^1_n)$ and
  $p^2 = \pi(i^2_1) \pi(i^2_2) \ldots \pi(i^2_n)$ be the two patterns
  from $\pi$ induced by $I^1$ and $I^2$, respectivelly.
  Clearly $p^1$ and $p^2$ are disjoint in $\pi$ (since $\mathcal{M}$ is
  a matching) and cover $\pi$ (since $\mathcal{M}$ is perfect).
  We also claim that $p^1$ and $p^2$ are order-isomorphic.
  Suppose, aiming at a contradiction, that this is not the case.
  Therefore, there exist two positions $j$ and $k$, say $j < k$, such that
  $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$, or
  $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$.
  But since $\mathcal{M}$ is containment-free, it follows that
  both $\{i^1_j, i^2_j\}$, $i^1_j < i^2_j$, and
  $\{i^1_k, i^2_k\}$, $i^1_k, i^2_k$, belong to $\mathcal{M}$.
  This is a contradiction since
  for any two disctinct edges
  $(a, b)$ and $(a', b')$ in $\mathcal{M}$
  we have $\pi(a) < \pi(a')$ if and only if $\pi(b) < \pi(b')$.
  \qed
\end{proof}

Let $\pi$ be a permutation.
For the sake of clarity, we will say that a bunch of consecutive positions $P$
of $\pi$ is \emph{above} (resp. \emph{below}) above another bunch of consecutive
positions $P'$ in permutation $\pi$
if $\pi(i) > \pi(j)$ (resp. $\pi(i) < \pi(j)$) for every
$i \in P$ and every $j \in P'$.
For exemple, $\sigma_1$ is above $\sigma_2$
(in an equivalent manner, $\sigma_2$ is below $\sigma_1$)
in Fig.~\ref{subfig:increasing before and aboce decreasing},
whereas $\sigma_1$ is below $\sigma_2$
(in an equivalent manner, $\sigma_2$ is above $\sigma_1$)
in Fig.~\ref{subfig:decreasing before and below increasing}.

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing and
    $\sigma_1$ is above $\sigma_2$
    (see Fig.~\ref{subfig:increasing before and aboce decreasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing and
    $\sigma_1$ is below $\sigma_2$
    (see Fig.~\ref{subfig:decreasing before and below increasing}),
    then $\mathcal{M}$ sontains at most one $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

>>>>

We denote by $S_n$ the set of permutations of size $n$ and by $S$ the
set of all permutations. In this paper, permutations of a size $n$ are
specified by words of length $n$ on the alphabet $[n]$ and without
multiple occurrence of a letter, so that all above definitions about
words remain valid on permutations. The only difference lies on the
fact that we shall denote by $\pi(i)$ (instead of $\pi_i$) the $i$-th
letter of any permutation $\pi$. For any nonnegative integer $n$, we
write $\nearrow_{n}$ (resp. $\searrow_{n}$) for the permutation
$1 2 \dots n$ (resp. $n (n-1) \dots 1$). If $\pi$ is a permutation of
$S_n$, we denote by $\bar \pi$ the {\em complementary} of $\pi$ that is
the permutation satisfying $\bar \pi(i) = n - \pi(i) + 1$ for all
$i \in [n]$. The {\em inverse} of $\pi$ is denoted by $\pi^{-1}$.

If $u$ is a word of integers without multiple occurrences of a same
letter, $\STD(u)$ is the {\em standarized} of $u$, that is the unique
permutation of the same size as $u$ such that for all $i, j \in [|u|]$,
$u_i < u_j$ if and only if $\STD(u)(i) < \STD(u)(j)$. In particular, the
image of the map $\STD$ is the set $S$ of all permutations. Two words $u$
and $v$ having the same standarized are {\em order-isomorphic}. If
$\sigma$ is a permutation, there is an {\em occurrence} of (the
{\em pattern}) $\sigma$ in $\pi$ if there is a set $P$ of indexes of
letters of $\pi$ such that $\sigma$ and $\pi_{|P}$ are order-isomorphic.
When $\pi$ does not admit any occurrence of $\sigma$, $\pi$ {\em avoids}
$\sigma$.

Let us now provide some definitions about graphs and matching, used in
the sequel. For a graph $G$, we denote $\VS{G}$ as the set of vertices
and $\ES{G}$ as the set of edges of $G$. Let $u = u_1 u_2 \dots u_n$ be
a word. The \emph{clique graph associated to $u$}, denoted by $K_u$, is
the complete graph on $\VS{K_u} = \{u_1, u_2, \ldots, u_n\}$. The
structure of this underlying graph is linear, \emph{i.e.}, the set of
its vertices is equipped with a natural total order $<$ defined by
$u_i < u_j$ if and only if $i < j$. For a linear graph, we write
$(u_i, u_j)$ instead of $\{u_i, u_j\}$ for an edge if it is clear from
the context that $i < j$. Recall that two edges of a graph are
\emph{independent} if they do not share a common vertex, and that a
\emph{matching} $\mathcal{M}$ in $G$ is a set of pairwise independent
edges. A matching is \emph{perfect} if it covers all the vertices of the
graph. In case the set of vertices is equipped with a total order, a
matching $\mathcal{M}$ is said to be \emph{containment-free} if there do
not exist (independent) edges $(u_i, u_j)$ and $(u_k, u_\ell)$ in
$\mathcal{M}$ such that $u_i < u_k < u_\ell < u_j$ or
$u_k < u_i < u_j < u_\ell$.

@@

        Finally, observe that any permutation $\pi$ avoiding $213$ and
        $231$ necessarily starts by the smallest possible letter or the
        greatest possible letter. This property is then true for the
        suffix of $\pi$ obtained by deleting its first letter,
        and so on for all of its suffixes. Thus, by
        replacing each letter $a$ of $\pi$ by $0$ (resp. $1$) if $a$ has the
        role of a smallest (resp. greatest) letter, one obtains a binary
        word $u$ such that $\BINTOPERM(u) = \pi$. Hence, all permutations
        avoiding $213$ and $231$ are in the image of $\BINTOPERM$.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_binary_to_square_permutation}
        If $u$ is a square binary word, $\BINTOPERM(u)$ is a square
        permutation.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_binary_to_square_permutation}]
        Since $u$ is a square binary word, there is a binary word $v$
        such that $u \in v \shuffle v$. Then, there are two disjoint
        sets $P$ and $Q$ of positions of letters of $u$ such that
        $u_{|P} = v = u_{|Q}$. Now, by definition of $\BINTOPERM$, the
        words $\BINTOPERM(u)_{|P}$ and $\BINTOPERM(u)_{|Q}$ have the
        same standarized $\sigma$. Hence, and by definition of
        the shuffle product of permutations, $\BINTOPERM(u)$ appears in
        $\sigma \SHUFFLE \sigma$, showing that $\BINTOPERM(u)$ is a
        square permutation.
        \qed
    \end{proof}
    \begin{claim} \label{claim:square_permutation_to_square_binary}
        If $\pi$ is a square permutation avoiding $213$ and $231$,
        $\PERMTOBIN(\pi)$ is a square binary word.
    \end{claim}
    \begin{proof}[of Claim~\ref{claim:square_permutation_to_square_binary}]
        Let $\pi$ be a square permutation avoiding $213$ and $231$. By
        Claim~\ref{claim:binary_to_permutation_avoiding}, $\pi$ is in
        the image of $\BINTOPERM$ and hence, $u = \PERMTOBIN(\pi)$ is a
        well-defined binary word. Since $\pi$ is a square permutation,
        there are two disjoint sets $P_1$ and $P_2$ of indexes of letters
        of $\pi$ such that $\pi_{|P_1}$ and $\pi_{|P_2}$ are
        order-isomorphic. This implies, by the definitions of $\BINTOPERM$
        and $\PERMTOBIN$, that $u_{|P_1} = u_{|P_2}$, showing that $u$
        is a square binary word.
        \qed
    \end{proof}
\end{proof}
\medskip

@@

\begin{proposition} \label{prop:square_stability}
    Let $\pi$ be a square permutation and $\sigma$ be a square root of
    $\pi$. Then,
    \begin{enumerate}[label={\it (\roman*)},fullwidth]
        \item \label{item:square_stability_1}
        the permutation $\widetilde{\pi}$ is a square and
        $\widetilde{\sigma}$ is one of its square roots;
        \item \label{item:square_stability_2}
        the permutation $\bar \pi$ is a square and $\bar \sigma$ is one of
        its square roots;
        \item \label{item:square_stability_3}
        the permutation $\pi^{-1}$ is a square and $\sigma^{-1}$ is one of
        its square roots.
    \end{enumerate}
\end{proposition}
\begin{proof}[of Proposition~\ref{prop:square_stability}]
    All statements~\ref{item:square_stability_1},
    \ref{item:square_stability_2}, and~\ref{item:square_stability_3} are
    consequences of Lemma~\ref{lem:endomorphisms}. Indeed,
    since $\pi$ is a square permutation and $\sigma$ is a square root of
    $\pi$, by definition, $\pi$ appears in the product
    $\sigma \SHUFFLE \sigma$. Now, by Lemma~\ref{lem:endomorphisms},
    for any $j = 1, 2, 3$, since $\phi_j$ is a morphism of associative
    algebras from $\QQ[S]$ to $\QQ[S]$, $\phi_j$ commutes with the
    shuffle product of permutations $\SHUFFLE$. Hence, in particular,
    one has
    \begin{equation}
        \phi_j(\sigma \SHUFFLE \sigma) =
        \phi_j(\sigma) \SHUFFLE \phi_j(\sigma).
    \end{equation}
    Then, since $\pi$ appears in $\sigma \SHUFFLE \sigma$, $\phi_j(\pi)$
    appears in $\phi_j(\sigma \SHUFFLE \sigma)$ and appears also in
    $\phi_j(\sigma) \SHUFFLE \phi_j(\sigma)$. This shows that
    $\phi_j(\sigma)$ is a square root of $\phi_j(\pi)$ and
    implies~\ref{item:square_stability_1}, \ref{item:square_stability_2},
    and~\ref{item:square_stability_3}.
    \qed
\end{proof}
\medskip

@@

This section is devoted to proving hardness of recognizing square
permutations. In the same way as happens with words, we shall use a
linear graph framework where deciding whether a permutation is a square
reduces to computing some specific matching in the associated linear
graph~\cite{Buss:Soltys:2014,Rizzi:Vialette:CSR:2013}. It reads as follows.

\begin{lemma}
  \label{lemma:matching}
  Let $\pi$ be a permutation. The following statements are equivalent:
  \begin{enumerate}
    \item The permutation $\pi$ is a square.
    \item There exists a perfect containment-free matching $\mathcal{M}$
    of $G_\pi$ such that for any two distinct edges $(a, b)$ and
    $(a', b')$ in $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only
    if $\pi(b) < \pi(b')$.
  \end{enumerate}
\end{lemma}

@@

    (\emph{1} $\Rightarrow$ \emph{2}).
    Since $\pi$ is a square, it admits a square root $\sigma$. Then, by
    definition, there are two sets $I^1 = \{i^1_1 < i^1_2 < \dots < i^1_n\}$
    and $I^2 = \{i^2_1 < i^2_2 < \dots < i^2_n\}$ of disjoint indexes of
    letters of $\pi$ such that $\pi_{|I^1}$ and $\pi_{|I^2}$ are both
    order-isomorphic to $\sigma$. Let $2n = |\pi|$ (and hence
    $|\sigma| = n$). Let $G_\pi$ be the linear clique associated to $\pi$
    with vertices
    $\VS{G_\pi} = \{i^1_j : j \in [n]\} \cup \{i^2_j : j \in [n]\}$. It
    is easily seen that $\mathcal{M} = \{\{i^1_j, i^2_j\} : j \in [n]\}$
    is a subset of $\ES{G_\pi}$. Furthermore, $\mathcal{M}$ is a perfect
    matching since $I^1 \cap I^2 = \emptyset$, and
    $I^1 \cup I^2 = \{1, 2,  \dots 2n\}$. It is also containment-free.
    Indeed, if it is not the case then there would exist two edges
    $e = \{i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$, $j < k$, in
    $\mathcal{M}$ such that $i^1_j < i^1_k < i^2_k < i^2_j$ or
    $i^1_j < i^2_k < i^1_k < i^2_j$. This is a contradiction since
    $i^2_k > i^2_j$ if $k > j$. Finally, for any two distinct edges
    $e = \{\i^1_j, i^2_j\}$ and $e' = \{i^1_k, i^2_k\}$ of $\mathcal{M}$,
    we have $\pi(i^1_j) < \pi(i^2_j)$ if and only if
    $\pi(i^1_k) < \pi(i^2_k)$ since we are comparing in both cases two
    elements (at position $j$ and $k$) in two patterns that are
    order-isomorphic to $\sigma$.

    (\emph{2} $\Rightarrow$ \emph{1}).
    Let $I^1 = \{i^1_j : \exists \{i_j, i_k\} \in \mathcal{M}
    \;\text{with}\; j < k\}$, and
    $I^2 = \{i^2_k : \exists \{i_j, i_k\} \in \mathcal{M}
    \;\text{with}\; j < k\}$.
    Informally, $I^1$ (resp. $I^2$) is the set of the left (resp. right)
    endpoints of the edges in $\mathcal{M}$ assuming the positions
    $1, 2, \dots, 2$ are drawn along an horizontal line.
    Without loss of generality, assume $i^1_1 < i^1_2 < \dots < i^1_n$.
    % and $I^2 = \{i^2_1, i^2_2, \dots, i^2_n\}$.
    Let $p^1 = \pi(i^1_1) \pi(i^1_2) \dots \pi(i^1_n)$ and
    $p^2 = \pi(i^2_1) \pi(i^2_2) \dots \pi(i^2_n)$ be the two patterns
    from $\pi$ induced by $I^1$ and $I^2$, respectively. Clearly $p^1$
    and $p^2$ are disjoint in $\pi$ (since $\mathcal{M}$ is a matching)
    and cover $\pi$ (since $\mathcal{M}$ is perfect). We also claim that
    $p^1$ and $p^2$ are order-isomorphic. Suppose, aiming at a
    contradiction, that this is not the case. Therefore, there exist two
    positions $j$ and $k$, say $j < k$, such that
    $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$, or
    $\pi(i^1_j) < \pi(i^2_j)$ and $\pi(i^1_k) > \pi(i^2_k)$.
    But since $\mathcal{M}$ is containment-free, it follows that both
    $\{i^1_j, i^2_j\}$, $i^1_j < i^2_j$, and $\{i^1_k, i^2_k\}$,
    $i^1_k, i^2_k$, belong to $\mathcal{M}$. This is a contradiction
    since for any two distinct edges $(a, b)$ and $(a', b')$ in
    $\mathcal{M}$ we have $\pi(a) < \pi(a')$ if and only if
    $\pi(b) < \pi(b')$.
    \qed
\end{proof}

Let $\pi$ be a permutation. For the sake of clarity, we will say that a
bunch of consecutive positions $P$ of $\pi$ is \emph{above} (resp.
\emph{below}) above another bunch of consecutive positions $P'$ in $\pi$
if $\pi(i) > \pi(j)$ (resp. $\pi(i) < \pi(j)$) for every $i \in P$ and
every $j \in P'$. For example, $\sigma_1$ is above $\sigma_2$ (in an
equivalent manner, $\sigma_2$ is below $\sigma_1$) in
Fig.~\ref{subfig:increasing before and above decreasing}, whereas
$\sigma_1$ is below $\sigma_2$ (in an equivalent manner, $\sigma_2$ is
above $\sigma_1$) in Fig.~\ref{subfig:decreasing before and below
increasing}.

@@

\begin{lemma}
  \label{lemma:at most one edge monotone}
  Let $\pi = \pi_1 \, \sigma_1 \, \pi_2 \, \sigma_2 \, \pi_3$
  be a permutation with $|\sigma_1| \geq 2$ and $|\sigma_2| \geq 2$,
  and $\mathcal{M}$ be a containment-free perfect matching of $G_\pi$.
  The following assertions hold:
  \begin{enumerate}
    \item
    If $\sigma_1$ is increasing, $\sigma_2$ is decreasing, and
    $\sigma_1$ is above $\sigma_2$ (see Fig.~\ref{subfig:increasing before
    and above decreasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
    \item
    If $\sigma_1$ is decreasing, $\sigma_2$ is increasing, and $\sigma_1$
    is below $\sigma_2$ (see Fig.~\ref{subfig:decreasing before and below
    increasing}), then $\mathcal{M}$ contains at most one
    $(\sigma_1, \sigma_2)$-edge.
  \end{enumerate}
\end{lemma}

\begin{proof}[of Lemma~\ref{lemma:at most one edge monotone}]
  \begin{figure}[t!]
    \centering
    \subfigure[%
      An increasing pattern before and above a decreasing pattern.
    ]{%
      \begin{tikzpicture}
        [
          scale=0.325,
          label/.style={anchor=base}
        ]
        % increasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,5.8) grid (5.2,11.2);
        \foreach \x/\y in {0/6,1/7,2/8,3/9,4/10,5/11} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % decreasing pattern
        \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,-0.2) grid (11.2,5.2);
        \foreach \x/\y in {6/5,7/4,8/3,9/2,10/1,11/0} {
            \draw [fill=black] (\x,\y) circle (0.1);
        }
        % labels
        \node [label] (a) at (1,5) {$a$};
        \node [label] (b) at (4,5) {$b$};
        \node [label] (ap) at (7,-1) {$a'$};
        \node [label] (bp) at (10,-1) {$b'$};
        \node (nu1) at (2.5,4) {$\sigma_1$};
        \node (nu2) at (8.5,-2) {$\sigma_2$};
        % crossings edges
        \draw [line width=1pt]
        (1,7) .. controls +(0,12) and +(0,10) .. (7,4);
        \draw [line width=1pt]
        (4,10) .. controls +(0,12) and +(0,10) .. (10,1);
      \end{tikzpicture}
      \label{subfig:increasing before and above decreasing}
    }% end subfigure
    \qquad\qquad\qquad
    \subfigure[%
      A decreasing pattern before and below an increasing pattern.
    ]{%
    \begin{tikzpicture}
      [
        scale=0.325,
        label/.style={anchor=base}
      ]
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (-0.2,-0.2) grid (5.2,5.2);
      \foreach \x/\y in {0/5,1/4,2/3,3/2,4/1,5/0} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % decreasing pattern
      \draw[step=1cm,black!30,ultra thin,fill=black!10] (5.8,5.8) grid (11.2,11.2);
      \foreach \x/\y in {6/6,7/7,8/8,9/9,10/10,11/11} {
          \draw [fill=black] (\x,\y) circle (0.1);
      }
      % labels
      \node [label] (a) at (1,-1) {$a$};
      \node [label] (b) at (4,-1) {$b$};
      \node [label] (ap) at (7,5) {$a'$};
      \node [label] (bp) at (10,5) {$b'$};
      \node (nu1) at (2.5,-2) {$\sigma_1$};
      \node (nu2) at (8.5,4) {$\sigma_2$};
      % crossings edges
      \draw [line width=1pt]
      (1,4) .. controls +(0,10) and +(0,12) .. (7,7);
      \draw [line width=1pt]
      (4,1) .. controls +(0,10) and +(0,12) .. (10,10);
      \end{tikzpicture}
      \label{subfig:decreasing before and below increasing}
    }% end subfigure
    \caption{\label{fig:subfig:no (nu_1, nu_2)-edge}%
      Illustration of Lemma~\ref{lemma:at most one edge monotone}.
    }
  \end{figure}
  We first prove \emph{1}. Suppose, aiming at a contradiction, that
  there exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$
  in $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may
  assume $a < b$ and $a' < b'$. Therefore, we have $\pi(a) < \pi(b)$
  (since $\sigma_1$ is increasing) and $\pi(a') > \pi(b')$ (since
  $\sigma_2$ is decreasing), which contradicts our hypothesis about
  $\mathcal{M}$.

  As for \emph{2}, suppose, still aiming at a contradiction, that there
  exist two $(\sigma_1, \sigma_2)$-edges $(a, a')$ and $(b, b')$ in
  $\mathcal{M}$. Since $\mathcal{M}$ is containment-free, we may assume
  $a < b$ and $a' < b'$. Therefore, we have $\pi(a) > \pi(b)$ (since
  $\sigma_1$ is decreasing) and $\pi(a') < \pi(b')$ (since $\sigma_2$ is
  increasing), which contradicts our hypothesis about~$\mathcal{M}$.
  \qed
\end{proof}
